Backward Population Synthesis: Mapping the Evolutionary History of Gravitational-Wave Progenitors

### 2022-10-19 src/tex/ms.tex
Fix L497
sorted
Closing out minor comments.

\begin{abstract}
One promising way to extract information about stellar astrophysics from
gravitational wave catalogs is to compare the catalog to the outputs of stellar
population synthesis modeling with varying physical assumptions.  The parameter
space of physical assumptions in population synthesis is high-dimensional and
the choice of parameters that best represents the evolution of a binary system
may depend in an as-yet-to-be-determined way on the system's properties.  Here
we propose a pipeline to simultaneously infer zero-age main sequence properties
and population synthesis parameter settings controlling modeled binary evolution
from individual gravitational wave observations of merging compact binaries. Our
pipeline can efficiently explore the high-dimensional space of population
synthesis settings and progenitor system properties for each system in a catalog
of gravitational wave observations.  We apply our pipeline to observations in
the third LIGO--Virgo Gravitational-Wave Transient Catalog. We showcase the
effectiveness of this pipeline with a detailed study of the progenitor
properties and population synthesis settings\textbf{, which represent 10 degrees
of freedom in total,} that produce mergers like the observed GW150914.  Our
pipeline permits a measurement of the variation of population synthesis
parameter settings with binary properties, if any; we \textbf{show that} the
recent GWTC-3 transient catalog \textbf{could be consistent with a model where
the stable mass transfer efficiency parameter varies with the primary black hole
mass.}  
\end{abstract}

@@

%One of the most common skepticisms of population synthesis studies is that they employ simulation sets which often
%cover a small subdomain of the entire parameter space of uncertain physics, or only vary one uncertain process per simulation
%(e.g. common envelope efficiency) while holding all others fixed to some ``default" values.
%On top of this, these values are often not calibrated to data.
Attempting to approach such an improved inference procedure for evolutionary
hyperparameters using traditional en mass Monte Carlo simulations of entire
populations of DCO mergers is computationally infeasible. Even with recent
developments in using emulators to speed up the simulation process
\citep[e.g.][]{Wong2021}, training the emulators still requires a significant
amount of computation to cover a wide range of uncertainty in the hyperparemeter
space. As an example, to emulate a model with $10$ parameters, the simplest way
to construct a training set of simulations is to run simulations on a grid which
varies combinations of uncertain physics. For $10$ uncertain physical processes,
if we consider $2$ variations for each physical process this would still require
$1024$ simulations. Because population synthesis simulations require hundreds to
thousands of cpu hours, training an emulator which spans a high dimensional
uncertainty space remains an impossibility at present.  Thus \textbf{it is
beneficial to infer hyperpameter settings \emph{dynamically}}, generating trial
DCO systems via population synthesis recipes to match a particular observation
one- or few-at-a-time while simultaneously adjusting the hyperparameters.
(Eventually it may be even be possible to use more physically-motivated modeling
of binary physics \citep[e.g.][]{Gallego-Garcia2021} in such a procedure.)

@@

In principle, we should only accept solutions that exactly reproduce the LVK posterior samples in the GW-observable space.
However, it is not feasible to achieve such a condition in practice, therefore we relax the condition in eq.~\ref{eq:loss}
to a small acceptance threshold. For this study, we picked a threshold of $10^{-2}$. To make sure we find a reasonably
complete set of progenitor parameters that corresponds to the posterior sample point, we use 1000 different initial guesses
in solving the optimization problem. As long as the solution fulfills the acceptance criteria, the root is as valid as all
other roots that fulfill the same acceptance criteria, \textbf{\emph{regardless of the initial guess used in finding the root}}.
This means, we have the freedom to choose the set of initial guesses however we think might benefit the optimization.

@@

Retaining explicit control over random variables allows us to marginalize over
their contribution, and thus allows us to focus on progenitor parameters and
hyperparameters. The strength and direction of natal kicks for compact objects 
are directly specified in the \texttt{COMPAS} binary population synthesis code 
\citep{Riley2022} and can also be specified in \texttt{COSMIC}, however 
this requires the user to specify the strength and direction of natal kicks as 
inputs through a user-specified pre-processing script.  In this study, we use 
\texttt{COSMIC} in it's default state which does not specify the kick strength 
and direction at run time. %However, in practice most population synthesis codes (including
%\texttt{COSMIC} which we use in this study) define their random variables
%implicitly such that the random variables are drawn within the program instead
%of being passed as arguments by default 
This means that the process of evolving the binary is not fully
deterministic\textbf{, i.e. we only have access to $\bm{\theta'}$ and
$\bm{\lambda}$ but not $\bm{X}$}. Thus, even if the root-finding algorithm
performs perfectly, forward modelling a set of roots does not guarantee that the
simulated population reproduces exactly the set of posterior samples due to
randomness in the evolution of each binary. To assure that the recovered
progenitor and hyperparameters robustly correspond to the posterior in the
GW-observable space, we push forward (or ``reproject'') the recovered
evolutionary parameters to the GW-observable space to check whether the
reprojected posterior agrees with the posterior given by the LVK collaboration.
\textbf{In a symbolic form, this means we generate a new $\bm{X}$ for each pair of $\bm{\theta'}$ and
$\bm{\lambda}$, then compute the new posterior in $\bm{\theta}$.}

@@

% Paragraph discussing the potential application on the population side.
To illustrate the potential benefit of using our method on the population level,
we perform the same analysis for all events in GWTC-3.
For events announced in GWTC-1, we use \texttt{Overall\_posterior}, \texttt{PublicationSamples},\texttt{IMRPhenomXPHM\_comoving} and \texttt{C01:Mixed} posterior samples, respectively.
All the data are publicly available from the Gravitational wave Open Science Center \citep{LIGOScientific:2019lzm}.
Plotted in figure
\ref{fig:GWTC-3_f_acc_mass} is the posterior density in the $m_{1,{\rm
GW}}-f_{\rm acc}$ space for most of the events in GWTC-3.
Each contour represents
the $68\%$ credible interval of the posterior density for that particular event.
There is a suggestive trend showing that $f_{\rm acc}$ could increase as the mass of
the progenitor increases. Such a trend implies the stable mass transfer
phase of a less massive binary would be preferentially non-conservative, with
more conservative mass transfer in more massive binary systems.

@@

We present a new pathway to understand binary evolution with GW events in this
paper. Instead of forward modelling an assumed distribution of initial binaries
to the observed population, we find the corresponding evolutionary parameters
event-by-event. In this first work, we showcase the power of the proposed method
with an application to GW150914. We show the joint posterior of the event's
progenitor parameters and hyperparameters.  Our work is both a more efficient
way to study GW event progenitors, and also allows the possibility of
constraining the astrophysics related to binary evolution, especially by
capturing the correlation between hyperparameters in different systems.

@@

% Rate analysis is now just counting, like what's done in phenomenological model, instead of forward modelling.
Working event-by-event as we do permits post-processing application of
arbitrarily complicated models of star formation and metalliticy evolution by
re-weighting the samples we obtain \citep[e.g.][]{VanSon2022}. Once we have
pulled the GW event posterior back from the observable space to the evolutionary
parameter space, we can apply the same hierarchical inference methods currently
used to infer the compact binary population to our progenitor population.  In
particular, we can use our progenitor population to infer the formation rate of
compact binary population progenitor systems over cosmic time without needing to
make artifical and simplistic assumptions about the delay time distribution or
the metallicity-specific star formation rate
\citep{Vitale2019,Ng2021,VanSon2022}.  Comparing the formation rate of
progenitors to the star formation rate provides an avenue to check our
understanding of binary evolution and the relative rate of other formation
channels.

% Single formation channel and fundamental limitation on that.
While our method allows data-driven exploration of the hyperparameters space for the first time, there are a number of improvements that can be implemented in future studies.
In this study we use only \texttt{COSMIC} as our evolutionary function, which by design cannot explain all the events in GWTC-3.
For example, events with either of the component mass larger than the lower edge of PISN gap such as GW190521 cannot be explained by \texttt{COSMIC}.
Alternative channels such as dynamical formation might be needed to explain some subset of the events in GWTC-3 \citep{Zevin2021}.
As the sensitivity of GW detectors increases, we expect to see more and more events that are unusual in some way.
Therefore, having a self-consistence population synthesis code that contains multiple formation channels is essential to accommodate the growing catalog of GW events.
In this paper, our main focus is to illustrate the concept of "back-propagating" GW-event posterior samples,
highlighting the capability of our method and motivating the benefits of building population synthesis codes that can work seamlessly with our method.
To avoid cluttering of focus, we discuss the physical implications of the results presented in this work under our specific assumptions (i.e. using \texttt{COSMIC} as our evolutionary function) in a future companion paper.

@@

To summarize, we propose a novel method to recover the posterior samples in the evolutionary parameter space for each GW event.
We point out hyperparameters in the usual population synthesis simulation context are not actually parameters related to the population,
but parameters about the evolutionary function.
This means the binary evolution functions can be constrained on an individual event basis.
We "back propagate" the posterior in the observable space to the evolutionary parameter space,
thus allowing us to study hyperparameters and theirs correlations with progenitor parameters in a data-driven manner.
Our method makes less assumptions than the traditional forward modelling approach,
which often fix the hyperparameters across the entire population.
Since we are not limited to the fixed hyperparameters assumptions, we can explore the behavior of the hyperparameters across the population much more efficiently.
While our work lays down a data analysis pathway to understand the population of GW events,
no physics can be learned without a comprehensive physical model.
We hope this letter will motivate the construction of next-generation population synthesis codes that have the following properties:
first, they should retain explicit control over the random variables so marginalizing over random variables can be done precisely and
second, they should be as automatically differentiable as possible so that exploring the evolutionary parameter space is efficient.
By combining the methods presented in this work and future, differentiable population synthesis codes,
we can explore the full parameter space of binary evolution models with future GW data.

>>>>

\begin{abstract}
One promising way to extract information about stellar astrophysics from
gravitational wave catalogs is to compare the catalog to the outputs of stellar
population synthesis modeling with varying physical assumptions.  The parameter
space of physical assumptions in population synthesis is high-dimensional and
the choice of parameters that best represents the evolution of a binary system
may depend in an as-yet-to-be-determined way on the system's properties.  Here
we propose a pipeline to simultaneously infer zero-age main sequence properties
and population synthesis parameter settings controlling modeled binary evolution
from individual gravitational wave observations of merging compact binaries. Our
pipeline can efficiently explore the high-dimensional space of population
synthesis settings and progenitor system properties for each system in a catalog
of gravitational wave observations.  We apply our pipeline to observations in
the third LIGO--Virgo Gravitational-Wave Transient Catalog. We showcase the
effectiveness of this pipeline with a detailed study of the progenitor
properties and population synthesis settings that produce mergers like the
observed GW150914.  Our pipeline permits a measurement of the variation of
population synthesis parameter settings with binary properties, if any; we
present inferences for the recent GWTC-3 transient catalog that suggest that the
stable mass transfer efficiency parameter may vary with primary black hole mass.  
\end{abstract}

@@

%One of the most common skepticisms of population synthesis studies is that they employ simulation sets which often
%cover a small subdomain of the entire parameter space of uncertain physics, or only vary one uncertain process per simulation
%(e.g. common envelope efficiency) while holding all others fixed to some ``default" values.
%On top of this, these values are often not calibrated to data.
Attempting to approach such an improved inference procedure for evolutionary
hyperparameters using traditional en mass Monte Carlo simulations of entire
populations of DCO mergers is computationally infeasible. Even with recent
developments in using emulators to speed up the simulation process
\citep[e.g.][]{Wong2021}, training the emulators still requires a significant
amount of computation to cover a wide range of uncertainty in the hyperparemeter
space. As an example, to emulate a model with $10$ parameters, the simplest way
to construct a training set of simulations is to run simulations on a grid which
varies combinations of uncertain physics. For $10$ uncertain physical processes,
if we consider $2$ variations for each physical process this would still require
$1024$ simulations. Because population synthesis simulations require hundreds to
thousands of cpu hours, training an emulator which spans a high dimensional
uncertainty space remains an impossibility at present.  Thus inference of
hyperpameter settings must proceed \emph{dynamically}, generating trial DCO
systems via population synthesis recipes to match a particular observation one-
or few-at-a-time while simultaneously adjusting the hyperparameters. (Eventually
it may be even be possible to use more physically-motivated modeling of binary
physics \citep[e.g.][]{Gallego-Garcia2021} in such a procedure.)

@@

In principle, we should only accept solutions that exactly reproduce the LVK posterior samples in the GW-observable space.
However, it is not feasible to achieve such a condition in practice, therefore we relax the condition in eq.~\ref{eq:loss}
to a small acceptance threshold. For this study, we picked a threshold of $10^{-2}$. To make sure we find a reasonably
complete set of progenitor parameters that corresponds to the posterior sample point, we use 1000 different initial guesses
in solving the optimization problem. As long as the solution fulfills the acceptance criteria, the root is as valid as all
other roots that fulfill the same acceptance criteria, \emph{regardless to its initial guess}.
This means, we have the freedom to choose the set of initial guesses however we think might benefit the optimization.

@@

Retaining explicit control over random variables allows us to marginalize over
their contribution, and thus allows us to focus on progenitor parameters and
hyperparameters. The strength and direction of natal kicks for compact objects 
are directly specified in the \texttt{COMPAS} binary population synthesis code 
\citep{Riley2022} and can also be specified in \texttt{COSMIC}, however 
this requires the user to specify the strength and direction of natal kicks as 
inputs through a user-specified pre-processing script.  In this study, we use 
\texttt{COSMIC} in it's default state which does not specify the kick strength 
and direction at run time. %However, in practice most population synthesis codes (including
%\texttt{COSMIC} which we use in this study) define their random variables
%implicitly such that the random variables are drawn within the program instead
%of being passed as arguments by default 
This means that the process of evolving the binary
is not fully deterministic. Thus, even if the root-finding algorithm performs
perfectly, forward modelling a set of roots does not guarantee that the
simulated population reproduces exactly the set of posterior samples due to
randomness in the evolution of each binary. To assure that the recovered
progenitor and hyperparameters robustly correspond to the posterior in the
GW-observable space, we push forward (or ``reproject'') the recovered
evolutionary parameters to the GW-observable space to check whether the
reprojected posterior agrees with the posterior given by the LVK collaboration.

@@

% Paragraph discussing the potential application on the population side.
To illustrate the potential benefit of using our method on the population level,
we perform the same analysis for all events in GWTC-3. For events announced in
GWTC-1, we use \texttt{Overall\_posterior},
\texttt{PublicationSamples},\texttt{IMRPhenomXPHM\_comoving} and
\texttt{C01:Mixed} posterior samples, respectively. All the data are publicly
available from the Gravitational wave Open Science Center
\citep{LIGOScientific:2019lzm}. Plotted in figure \ref{fig:GWTC-3_f_acc_mass} is
the posterior density in the $m_{1,{\rm GW}}-f_{\rm acc}$ space for
most\footnote{\textbf{We excluded events for which our reconstructed posterior has KL
divergence greater than 0.1 nats relative to the GWTC-3 posterior; see below.}}
of the events in GWTC-3. Each contour represents the $68\%$ credible interval of
the posterior density for that particular event. There is a suggestive trend
showing that $f_{\rm acc}$ could increase as the mass of the progenitor
increases. Such a trend implies the stable mass transfer phase of a less massive
binary would be preferentially non-conservative, with more conservative mass
transfer in more massive binary systems.

@@

We present a new pathway to understand binary evolution with GW events in this
paper. Instead of forward modelling an assumed distribution of initial binaries
to the observed population, we find the corresponding evolutionary parameters
event-by-event. In this first work, we showcase the power of the proposed method
with an application to GW150914. We show the joint posterior of the event's
progenitor parameters and hyperparameters.  Our work is a \textbf{data-driven}
way to study GW event progenitors, and also allows the possibility of
constraining the astrophysics related to binary evolution, especially by
capturing the correlation between hyperparameters in different systems.

@@

% Rate analysis is now just counting, like what's done in phenomenological model, instead of forward modelling.
Working event-by-event as we do permits post-processing application of
arbitrarily complicated models of star formation and metalliticy evolution by
re-weighting the samples we obtain \citep[e.g.][]{VanSon2022}. Once we have
pulled the GW event posterior back from the observable space to the evolutionary
parameter space, \textbf{and have accounted for selection effects}, we can apply
the same hierarchical inference methods currently used to infer the compact
binary population to our progenitor population.  In particular, we can use our
progenitor population to infer the formation rate of compact binary population
progenitor systems over cosmic time without needing to make artifical and
simplistic assumptions about the delay time distribution or the
metallicity-specific star formation rate \citep{Vitale2019,Ng2021,VanSon2022}.
Comparing the formation rate of progenitors to the star formation rate provides
an avenue to check our understanding of binary evolution and the relative rate
of other formation channels.

% Single formation channel and fundamental limitation on that.
While our method allows data-driven exploration of the hyperparameters space, there are a number of improvements that can be implemented in future studies.
In this study we use only \texttt{COSMIC} as our evolutionary function, which by design cannot explain all the events in GWTC-3.
For example, events with either of the component mass larger than the lower edge of PISN gap such as GW190521 cannot be explained by \texttt{COSMIC}.
Alternative channels such as dynamical formation might be needed to explain some subset of the events in GWTC-3 \citep{Zevin2021}.
As the sensitivity of GW detectors increases, we expect to see more and more events that are unusual in some way.
Therefore, having a self-consistence population synthesis code that contains multiple formation channels is essential to accommodate the growing catalog of GW events.
In this paper, our main focus is to illustrate the concept of "back-propagating" GW-event posterior samples,
highlighting the capability of our method and motivating the benefits of building population synthesis codes that can work seamlessly with our method.
To avoid cluttering of focus, we discuss the physical implications of the results presented in this work under our specific assumptions (i.e. using \texttt{COSMIC} as our evolutionary function) in a future companion paper.

@@

To summarize, we propose a method to recover the posterior samples in the evolutionary parameter space for each GW event.
We point out hyperparameters in the usual population synthesis simulation context are not actually parameters related to the population,
but parameters about the evolutionary function.
This means the binary evolution functions can be constrained on an individual event basis.
We "back propagate" the posterior in the observable space to the evolutionary parameter space,
thus allowing us to study hyperparameters and theirs correlations with progenitor parameters in a data-driven manner.
Our method makes less assumptions than the traditional forward modelling approach,
which often fix the hyperparameters across the entire population.
Since we are not limited to the fixed hyperparameters assumptions, we can explore the behavior of the hyperparameters across the population much more efficiently.
While our work lays down a data analysis pathway to understand the population of GW events,
no physics can be learned without a comprehensive physical model.
We hope this letter will motivate the construction of next-generation population synthesis codes that have the following properties:
first, they should retain explicit control over the random variables so marginalizing over random variables can be done precisely and
second, they should be as automatically differentiable as possible so that exploring the evolutionary parameter space is efficient.
By combining the methods presented in this work and future, differentiable population synthesis codes,
we can explore the full parameter space of binary evolution models with future GW data.
